---
title: "Exercises Chapter 6"
section: "6.4"
Exercise: "2", "9", "14", "19"
author: "Monaliben Pansuriya (275672)",
        "Rahul Bhunatar (276222)",
        "Aiswarya Pulluttiparambil (276209)",
        "Vaishanavi Chidambaram (275776)"
date: "2024-01-07"
output: html_document
---

## Exercise 2

Those responsible for the FranSO study (Kemmler et al. (2017)) assumed for the
sample size calculation of the primary endpoint (sarcopenia z-score), a difference
of means of 1.0 and a standard deviation of 1.4. Calculate the sample size using
Student’s t-test. Use a significance level of 5% and a power of 80%. How much does
the sample size increase if you increase the power to 90%? The study ultimately
included 33 respectively, 34 subjects per group. How large is the power of the
study if you assume a sample size of 33 per group?

## Answer

To calculate the sample size using Student’s t-test, we can use the `power.t.test`
function. 

Given the assumptions of a difference of means of 1.0, a standard deviation of
1.4, a significance level of 5%, and a power of 80%, we can calculate the sample
size as follows:

We perform the case number calculation with δ = 1.0, α = 1.4, σ = 0.05 and 1 - β = 0.8

```{r}
# Calculate sample size for 80% power
power.t.test(delta = 1.0, sd = 1.4, sig.level = 0.05, power = 0.8)
```

In these calculation, a case number of 32 per group is sufficient to produce
the expected difference with a power of 80% and an type 1 error of 5%.

To calculate the sample size for 90% power, we can use the same function with the
updated power value:

We perform the case number calculation with δ = 1.0, α = 1.4, σ = 0.05 and 1 - β = 0.9

```{r}
# Calculate sample size for 90% power
power.t.test(delta = 1.0, sd = 1.4, sig.level = 0.05, power = 0.9)
```

In these calculation, a case number of 42 per group is sufficient to produce
the expected difference with a power of 90% and an type 1 error of 5%.

Next, to determine the power of the study with a sample size of 33 per group,
we can use the `power.t.test` function as follows:

We perform the case number calculation with δ = 1.0, α = 1.4, σ = 0.05 and n = 33

```{r}
# Calculate power for a sample size of 33 per group
power.t.test(n = 33, delta = 1.0, sd = 1.4, sig.level = 0.05)
```

In these calculation, we use a case number of n = 33. With complete data this
leads to a power of approx 81%.

By running these R commands, you can obtain the sample sizes and power calculations
based on the given parameters and study design.


## Exercise 9

## Exercise 14

## Exercise 19